Appendix 1: sdmTMB statistical model description

### 2022-01-05 doc/appendix-model.Rmd
EW edits to model-appendix
Mostly minor notation stuff

* $y_{\s,t}$ represents the response data at point $\s$ and time $t$;
* $\mu$ represents the mean;
* $f$ represents a link function (e.g., log or logit) and $f^{-1}$ represents its inverse;
* $\bm{X}_{1}$, $\bm{X}_{2}$, and $\bm{X}_{3}$ represent design matrices;
* $\bm{\beta}$ represents a vector of fixed-effect coefficients;
* $O_{\s,t}$ represents an offset: a covariate (usually log transformed) with a coefficient fixed at one;
* $\alpha_{g}$ represents random intercepts by group $g$, $\alpha_{g}\sim \mathrm{N}(0,\sigma^2_\alpha)$; 
<!-- should alpha_g be bold as in the equation above? -->
* $\gamma_{t}$ represents time-varying coefficients (a random walk), $\gamma_{t} \sim \mathrm{N}(\gamma_{t-1},\sigma^2_\gamma)$;
* $\zeta_{\bm{s}}$ represents spatially varying coefficients (a random field), $\zeta_{\bm{s}} \sim \mathrm{MVN}(\bm{0},\bm{\Sigma}_\zeta)$;
* $\omega_{\bm{s}}$ represents a random spatial component (a random field), $\omega_{\bm{s}} \sim \mathrm{MVN}(\bm{0},\bm{\Sigma}_\omega)$; and
* $\epsilon_{\bm{s},t}$ represents a spatiotemporal component (a random field), $\epsilon_{\bm{s},t} \sim \mathrm{MVN}(\bm{0},\bm{\Sigma}_{\epsilon})$.

@@

### Smoothers

Smoothers in sdmTMB are implemented with the same formula syntax familiar to mgcv users fitting GAMs (generalized additive models).
Smooths are implemented in the formula using `+ s(x)`, which implements a smooth from `mgcv::s()`. 
Within these smooths, the same syntax commonly used in `mgcv::s()` can be applied, e.g. 2-dimensional smooths may be constructed with `+ s(x, y)`; smooths can be specific to various factor levels, `+ s(x, by = group)`; smooths can vary according to a continuous variable, `+ s(x, by = x2)`; the basis function dimensions may be specified, e.g. `+ s(x, k = 4)`; and various types of splines may be constructed such as cyclic splines to model seasonality, e.g. `+ s(month, bs = "cc", k = 12)`.

@@

## Spatial random fields

Spatial random fields, $\omega_s$, are included if `spatial = 'on'` and omitted if `spatial = 'off'`.

$$
\begin{aligned}
\mu_{\bm{s},t} &= f^{-1} \left( \ldots + \omega_s + \ldots \right),\\
\boldsymbol{\omega} &\sim \operatorname{MVNormal} \left( \boldsymbol{0}, \boldsymbol{\Sigma}_\omega \right),\\ 
\end{aligned}
$$
The marginal standard deviation of $\boldsymbol{\omega}$ is indicated by `Spatial SD` in the model output or as `sigma_O` in the output of `sdmTMB::tidy(fit, "ran_pars")`.

@@

$$
\boldsymbol{\omega} \sim \mathrm{MVNormal}\left(\boldsymbol{0}, \sigma_\omega^2 \boldsymbol{Q}^{-1}_\omega\right)
$$
where $\boldsymbol{Q}_\omega$ is a sparse precision matrix and $\sigma_\omega^2$ is the marginal variance.

@@

### IID spatiotemporal random fields

IID spatiotemporal random fields (`spatiotemporal = 'iid'`) can be represented as

$$
\begin{aligned}
\mu_{\bm{s},t} &= f^{-1} \left( \ldots + \epsilon_{\s,t} + \ldots \right),\\
\boldsymbol{\epsilon}_t &\sim \operatorname{MVNormal} \left( \boldsymbol{0}, \boldsymbol{\Sigma}_{\epsilon} \right).
\end{aligned}
$$

where $\epsilon_{\s,t}$ represent random field deviations at point $\s$ and time $t$.
The random fields are assumed independent across time slices.

@@

$$
\boldsymbol{\epsilon}_t \sim \mathrm{MVNormal}\left(\boldsymbol{0}, \sigma_\epsilon^2 \boldsymbol{Q}^{-1}_\epsilon\right)
$$

@@

$$
\begin{aligned}
\mu_{\s,t} &= f^{-1} \left( \ldots + \delta_{\s,t} \ldots \right),\\
\boldsymbol{\delta}_{t=1} &\sim \operatorname{MVNormal} (\boldsymbol{0}, \boldsymbol{\Sigma}_{\epsilon}),\\
\boldsymbol{\delta}_{t>1} &= \rho \boldsymbol{\delta}_{t-1} + \sqrt{1 - \rho^2} \boldsymbol{\epsilon}_t,  \:
\boldsymbol{\epsilon}_t \sim \operatorname{MVNormal} \left(\boldsymbol{0}, \boldsymbol{\Sigma}_{\epsilon} \right),
\end{aligned}
$$
where $\rho$ is the correlation between subsequent spatiotemporal random fields. 
The $\rho \boldsymbol{\delta}_{t-1} + \sqrt{1 - \rho^2}$ term scales the spatiotemporal variance by the correlation such that it represents the steady-state marginal variance.
The correlation $\rho$ allows for mean-reverting spatiotemporal fields, and is constrained to be $-1 < \rho < 1$.
Internally, the parameter is estimated as `ar1_phi`, which is unconstrained. 
The parameter `ar1_phi` is transformed to $\rho$ with $\rho = 2 \left( \mathrm{logit}^{-1}(\texttt{ar1\_phi}) - 1 \right)$.

@@

$$
\begin{aligned}
\mu_{\s,t} &= f^{-1} \left( \ldots + \delta_{\s,t} + \ldots \right),\\
\boldsymbol{\delta}_{t=1} &\sim \operatorname{MVNormal} (\boldsymbol{0}, \boldsymbol{\Sigma}_{\epsilon}),\\
\boldsymbol{\delta}_{t>1} &= \boldsymbol{\delta}_{t-1} +  \boldsymbol{\epsilon}_{t-1},  \:
\boldsymbol{\epsilon}_{t-1} \sim \operatorname{MVNormal} \left(\boldsymbol{0}, \boldsymbol{\Sigma}_{\epsilon} \right),
\end{aligned}
$$

@@

where $\boldsymbol{\zeta}$ is a random field representing a spatially varying coefficient.
Usually, $\bm{X}_{3,t}$ would represent a predictor that is constant spatially for a given time $t$.
Currently, $\bm{X}_{3,t}$ is restricted to a single covariate, supplied by a one-sided formula to `spatial_varying`.
For example `spatial_varying = ~ 0 + x`, where `0` omits the intercept.

@@

## IID random or multi-level intercepts

$$
\begin{aligned}
  \mu_{\s,t} &= f^{-1} \left( \ldots + \alpha_{g} + \ldots \right),\\
  \alpha_g &\sim \operatorname{Normal} \left(0, \sigma_\alpha^2 \right),\\
\end{aligned}
$$

@@

$$
\operatorname{Binomial} \left(N, \mu \right)
$$
where $N$, is the size or number of trials, and $\mu$ is the probability of success for each trial.
If $N = 1$, the distribution becomes the Bernoulli distribution.
Internally, the distribution is parameterized as the [robust
version](https://kaskr.github.io/adcomp/group__R__style__distribution.html#gaecb5a18095a320b42e2d20c4b120f5f5) in TMB, which is numerically stable when probabilities approach 0 or 1.

Following the structure of `stats::glm()`, lme4, and glmmTMB, a binomial family can be specified in one of 4 ways:

@@

$$
\operatorname{Beta} \left(\mu \phi, (1 - \mu) \phi \right)
$$

where $\mu$ is the mean and $\phi$ is a precision parameter.
This parametrization follows @ferrari2004 and the betareg R package [@cribari-neto2010].
The variance is $\mu (1 - \mu) / (\phi + 1)$.

@@

$$
\operatorname{Gamma} \left( \phi, \frac{\mu}{\phi}  \right)
$$

where $\phi$ represents the Gamma shape and $\mu / \phi$ represents the scale. 
The variance is $\mu \cdot \phi$.

@@

TODO: did this get reversed above? Isn't it set up so the mean(x) is mu?

Code defined [within sdmTMB](https://github.com/pbs-assess/sdmTMB/blob/18a39eabc111e2179fa589f942c8820d87ad10df/src/utils.h#L47-L54) based on the TMB `dnorm()` normal density.

>>>>

* $y_{\s,t}$ represents the response data at point $\s$ and time $t$;
* $\mu$ represents the mean;
* $f$ represents a link function (e.g., log or logit) and $f^{-1}$ represents its inverse;
* $\bm{X}_{1}$, $\bm{X}_{2}$, and $\bm{X}_{3}$ represent design matrices;
* $\bm{\beta}$ represents a vector of fixed-effect coefficients;
* $O_{\s,t}$ represents an offset: a covariate (usually log transformed) with a coefficient fixed at one;
* $\alpha_{g}$ represents random intercepts by group $g$, $\alpha_{g}\sim \mathrm{N}(0,\sigma^2_\alpha)$; 
<!-- should alpha_g be bold as in the equation above? --><!-- EW: yes -- alpha_g should be bolded here -->
* $\gamma_{t}$ represents time-varying coefficients (a random walk), $\gamma_{t} \sim \mathrm{N}(\gamma_{t-1},\sigma^2_\gamma)$;
* $\zeta_{\bm{s}}$ represents spatially varying coefficients (a random field), $\zeta_{\bm{s}} \sim \mathrm{MVN}(\bm{0},\bm{\Sigma}_\zeta)$;
* $\omega_{\bm{s}}$ represents a random spatial component (a random field), $\omega_{\bm{s}} \sim \mathrm{MVN}(\bm{0},\bm{\Sigma}_\omega)$; and
* $\epsilon_{\bm{s},t}$ represents a spatiotemporal component (a random field), $\epsilon_{\bm{s},t} \sim \mathrm{MVN}(\bm{0},\bm{\Sigma}_{\epsilon})$.

@@

### Smoothers
<!-- EW: see this paragraph and below, should package names be enclosed in single quotes, like `sdmTMB()` above? Examples: mgcv, brms, sdmTMB throughout -->
<!-- EW: unrelated, I never tried to implement factor smooths. I assume they work, but it'd good to check eventually -->
Smoothers in sdmTMB are implemented with the same formula syntax familiar to mgcv users fitting GAMs (generalized additive models).
Smooths are implemented in the formula using `+ s(x)`, which implements a smooth from `mgcv::s()`. 
Within these smooths, the same syntax commonly used in `mgcv::s()` can be applied, e.g. 2-dimensional smooths may be constructed with `+ s(x, y)`; smooths can be specific to various factor levels, `+ s(x, by = group)`; smooths can vary according to a continuous variable, `+ s(x, by = x2)`; the basis function dimensions may be specified, e.g. `+ s(x, k = 4)`; and various types of splines may be constructed such as cyclic splines to model seasonality, e.g. `+ s(month, bs = "cc", k = 12)`.

@@

## Spatial random fields
<!-- EW: edited omega in this section to be bolded and subscripted with s, for consistency with above --> 
Spatial random fields, $\omega_{\bm{s}}$, are included if `spatial = 'on'` and omitted if `spatial = 'off'`.

$$
\begin{aligned}
\mu_{\bm{s},t} &= f^{-1} \left( \ldots + \boldsymbol{\omega_s} + \ldots \right),\\
\boldsymbol{\omega_s} &\sim \operatorname{MVNormal} \left( \boldsymbol{0}, \boldsymbol{\Sigma}_\omega \right),\\ 
\end{aligned}
$$
The marginal standard deviation of $\boldsymbol{\omega}$ is indicated by `Spatial SD` in the model output or as `sigma_O` in the output of `sdmTMB::tidy(fit, "ran_pars")`.

@@

$$
\boldsymbol{\omega_s} \sim \mathrm{MVNormal}\left(\boldsymbol{0}, \sigma_\omega^2 \boldsymbol{Q}^{-1}_\omega\right)
$$
where $\boldsymbol{Q}_\omega$ is a sparse precision matrix and $\sigma_\omega^2$ is the marginal variance.

@@

### IID spatiotemporal random fields
<-- EW: as above, I changed epsilon to be subscripted with s throughout this section-->
IID spatiotemporal random fields (`spatiotemporal = 'iid'`) can be represented as

$$
\begin{aligned}
\mu_{\bm{s},t} &= f^{-1} \left( \ldots + \boldsymbol{\epsilon_{\s,t}} + \ldots \right),\\
\boldsymbol{\epsilon_{\s,t}} &\sim \operatorname{MVNormal} \left( \boldsymbol{0}, \boldsymbol{\Sigma}_{\epsilon} \right).
\end{aligned}
$$

where $\boldsymbol{\epsilon_{\s,t}}$ represent random field deviations at point $\s$ and time $t$.
The random fields are assumed independent across time slices.

@@

$$
\boldsymbol{\epsilon_{\s,t}} \sim \mathrm{MVNormal}\left(\boldsymbol{0}, \sigma_\epsilon^2 \boldsymbol{Q}^{-1}_\epsilon\right)
$$

@@

$$
\begin{aligned}
\mu_{\s,t} &= f^{-1} \left( \ldots + \delta_{\s,t} \ldots \right),\\
\boldsymbol{\delta}_{t=1} &\sim \operatorname{MVNormal} (\boldsymbol{0}, \boldsymbol{\Sigma}_{\epsilon}),\\
\boldsymbol{\delta}_{t>1} &= \rho \boldsymbol{\delta}_{t-1} + \sqrt{1 - \rho^2} \boldsymbol{\epsilon_{\s,t}},  \:
\boldsymbol{\epsilon_{\s,t}} \sim \operatorname{MVNormal} \left(\boldsymbol{0}, \boldsymbol{\Sigma}_{\epsilon} \right),
\end{aligned}
$$
where $\rho$ is the correlation between subsequent spatiotemporal random fields. 
The $\rho \boldsymbol{\delta}_{t-1} + \sqrt{1 - \rho^2}$ term scales the spatiotemporal variance by the correlation such that it represents the steady-state marginal variance.
The correlation $\rho$ allows for mean-reverting spatiotemporal fields, and is constrained to be $-1 < \rho < 1$.
Internally, the parameter is estimated as `ar1_phi`, which is unconstrained. 
The parameter `ar1_phi` is transformed to $\rho$ with $\rho = 2 \left( \mathrm{logit}^{-1}(\texttt{ar1\_phi}) - 1 \right)$.

@@

$$
\begin{aligned}
\mu_{\s,t} &= f^{-1} \left( \ldots + \delta_{\s,t} + \ldots \right),\\
\boldsymbol{\delta}_{t=1} &\sim \operatorname{MVNormal} (\boldsymbol{0}, \boldsymbol{\Sigma}_{\epsilon}),\\
\boldsymbol{\delta}_{t>1} &= \boldsymbol{\delta}_{t-1} +  \boldsymbol{\epsilon_{\s,t-1}},  \:
\boldsymbol{\epsilon_{\s,t-1}} \sim \operatorname{MVNormal} \left(\boldsymbol{0}, \boldsymbol{\Sigma}_{\epsilon} \right),
\end{aligned}
$$

@@

where $\boldsymbol{\zeta}$ is a random field representing a spatially varying coefficient.
Usually, $\bm{X}_{3,t}$ would represent a predictor that is constant spatially for a given time $t$.
Currently, $\bm{X}_{3,t}$ is restricted to a single covariate (column vector), supplied by a one-sided formula to `spatial_varying`.
For example `spatial_varying = ~ 0 + x`, where `0` omits the intercept.

@@

## IID random or multi-level intercepts
<-- EW: make alpha bold here, as above?-->
$$
\begin{aligned}
  \mu_{\s,t} &= f^{-1} \left( \ldots + \alpha_{g} + \ldots \right),\\
  \alpha_g &\sim \operatorname{Normal} \left(0, \sigma_\alpha^2 \right),\\
\end{aligned}
$$

@@

$$
\operatorname{Binomial} \left(N, \mu \right)
$$
where $N$, is the size or number of trials, and $\mu$ is the probability of success for each trial.
If $N = 1$, the distribution becomes the Bernoulli distribution.
Internally, the distribution is parameterized as the [robust
version](https://kaskr.github.io/adcomp/group__R__style__distribution.html#gaecb5a18095a320b42e2d20c4b120f5f5) in TMB, which is numerically stable when probabilities approach 0 or 1.
<-- EW: Put lme4 and glmmTMB in '' too?-->
Following the structure of `stats::glm()`, lme4, and glmmTMB, a binomial family can be specified in one of 4 ways:

@@

$$
\operatorname{Beta} \left(\mu \phi, (1 - \mu) \phi \right)
$$
<<- EW: betareg in ''?-->
where $\mu$ is the mean and $\phi$ is a precision parameter.
This parametrization follows @ferrari2004 and the betareg R package [@cribari-neto2010].
The variance is $\mu (1 - \mu) / (\phi + 1)$.

@@

$$
\operatorname{Gamma} \left( \phi, \frac{\mu}{\phi}  \right)
$$
<-- EW: check this, I think the Gamma variance should be mu * phi2 -->
where $\phi$ represents the Gamma shape and $\mu / \phi$ represents the scale. 
The mean is $\mu$ and variance is $\mu \cdot \phi^2$.

@@

TODO: did this get reversed above? Isn't it set up so the mean(x) is mu?
<-- EW: the lognormal definition above includes the bias correction phi2/2, but that's not in the standard implementations in R, etc. I think I used those R functions or referenced Wikipedia in writing the equations out. So we should pick one or the other. I think just changing the definition to not include the bias correction would keep it consistent with R, https://stat.ethz.ch/R-manual/R-devel/library/stats/html/Lognormal.html-->
Code defined [within sdmTMB](https://github.com/pbs-assess/sdmTMB/blob/18a39eabc111e2179fa589f942c8820d87ad10df/src/utils.h#L47-L54) based on the TMB `dnorm()` normal density.
